Supplement: Autopolyploid establishment through polygenic adaptation

### 2025-03-24 doc/paper/supp.tex
starting revisions
update

\begin{document}
\title{Supplement: Autopolyploid establishment through polygenic adaptation}
\author{Arthur Zwaenepoel}
\date{\vspace{-5ex}}
\maketitle
\tableofcontents
\clearpage

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr-check.pdf}
\caption{
    Nonzero $\alpha_k$ parameters increase the tetraploid establishment
    probability, but mainly due to $\alpha_2$, which increases the segregation
    variance associated with the formatoin of diploid gametes by diploids.
    All other parameters and simulation details are as in \cref{m-fig:estwmig}.
\label{fig:suppdr}}
\end{figure}

@@

For finite $N$, the basic mixed-ploidy model defines a Markov chain on the
state space $[0..N] \times [0..N]$.
\begin{align}
  p_{ij,kl} &= \Pr\{N_2(t+1)=k, N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
    &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
\end{align}
where 
\begin{align}
  p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
  p_3 &= \left(\frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
    {N(1-u) + (i+j)u + j(2v-1)}\right)^2 
\end{align}
Associating a unique index with each pair $(i,j)$ with $0 \le i,j \le N$, we
can define a transition probability matrix $P$ of dimensions $(N+1)^2 \times
(N+1)^2$ for this Markov chain.

@@

There is little difficulty in extending the recursions for diploids
\citep{barton2017} and autotetraploids (\cref{eq:rectet}) to the mixed-ploidy
case. 
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$, as in the
autotetraploid model.
The recursion for the coancestry coefficients in is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

@@

Consider a single locus in a population of $k$-ploids, and let $X_1,\dots,X_k$
denote a random genotype at this locus, where $X_i$ is the additive genetic
value of the allele on homolog $i$.
The variance in gametic values $Y$ produced by some meiotic process can be
decomposed as 
\begin{equation}
  \Var[Y] = \underbrace{\Ex\left[\Var[Y|X_1, \dots, X_k]\right]}_{\text{segregation variance}} 
    + \underbrace{\Var\left[\Ex[Y|X_1, \dots, X_k]\right]}_{\text{population
variation}}
\label{eq:eve}
\end{equation}
We use this relationship to derive the segregation variance for a given meiotic
process.

As an example, consider ordinary diploid meiosis at a single locus, where a
diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
gene from the population and $v_x$ is the variance in additive genetic values
at the locus.
We have
  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
Using \cref{eq:eve}, we can then find the segregation variance
  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
Summing over $L$ independent loci, we find the gametic segregation variance in
diploids (which is halve the zygotic segregation variance $V$, by definition) as
  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{\Vx_{2}}{2}$$
and hence $V=\Vx_{2}$, where $\Vx_{2}$ is the genetic variance associated with
a haploid genome in the diploid cytotype, as defined in the main text.
This will hold in the infinitsimal limit, where $L$ becomes very large and
$v_x$ smaller and smaller.

Below, we derive the segregation variance associated with haploid and diploid
gamete production in the three cytotypes of the mixed-ploidy population.
We use \cref{eq:eve} to do this, and express the segregation variance for a
$k$-ploid individual producing an $l$-ploid $V_{S(k,l)}$ in terms of the
variance $\Vx_{k}$ associated with a haploid genome in a (hypothetical)
$k$-ploid reference population at HWLE. Using the scaling assumptions outlined
in the main text, i.e. $\Vx_{k} = \beta_k^2 V$, we can then express all
segregation variances in terms of the diploid segregation variance in the
reference population $V$.

@@

The mechanisms of unreduced gamete formation do not necessarily lead to a
faithful transmission of the complete diploid genome.
Unreduced gametes are formed in two ways, depending on the meiotic abberation
that leads to their origin: (1) first division restitution (FDR) of (2) second
division restitution (SDR) \citep{bretagnolle1995,storme2013}.
Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
recombination happens with probability $c$ and that conditional on unreduced
gamete formation, formation is due to FDR with probability $f$ while it is due
to SDR with probability $1-f$.
The different unreduced gametes that are formed are represented schematically
in \cref{fig:udip}.
Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
locus will be
$$Y = \begin{cases}
    2X_1 & \text{w.pr.  } \alpha_2/2 \\
    2X_2 & \text{w.pr.  } \alpha_2/2 \\
    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
    \end{cases}$$
where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
of the same gene end up in a diploid gamete produced by a diploid individual
(see the diagram in \cref{fig:udip}). 
Conditional on the latter event, we get
\begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*}
Conditioning on the complementary event, all gametes have genetic value
$Y = X_1+X_2$, so that the segregation variance is 0.
Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 \Vx_{2}$.

>>>>

\begin{document}
\title{Supplement: Autopolyploid establishment through polygenic adaptation}
%\author{Arthur Zwaenepoel}
\author{}
\date{\vspace{-5ex}}
\maketitle
\tableofcontents
\clearpage

@@

\begin{figure}
\includegraphics[width=0.9\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr-check.pdf}
\caption{
    Nonzero $\alpha_k$ parameters increase the tetraploid establishment
    probability, but mainly due to $\alpha_2$, which increases the segregation
    variance associated with the formation of diploid gametes by diploids.
    All other parameters and simulation details are as in \cref{m-fig:estwmig}.
\label{fig:suppdr}}
\end{figure}

@@

For finite $N$, the basic mixed-ploidy model defines a Markov chain on the
state space $[0..N] \times [0..N]$.
\begin{align}
  p_{ij,kl} &= \Pr\{N_2(t+1)=k, N_3(t+1)=l | N_2(t)=i, N_3(t)=j\} \nonumber \\
    &= \frac{N!}{k!l!(N-k-l)!} p_2^k p_3^l (1-p_2-p_3)^{N-k-l}
\end{align}
where 
\begin{align}
  p_2 &= \left(\frac{i(1-u) + jv}{N(1-u) + (i+j)u + j(2v-1)}\right)^2 \\
  p_3 &= \frac{2(i(1-u) + jv)(N(1-u) + i(2u - 1) + j(u + v - 1))}
    {\left(N(1-u) + (i+j)u + j(2v-1)\right)^2} 
\end{align}
Associating a unique index with each pair $(i,j)$ with $0 \le i,j \le N$, we
can define a transition probability matrix $P$ of dimensions $(N+1)^2 \times
(N+1)^2$ for this Markov chain.

@@

There is little difficulty in extending the recursions for diploids
\citep{barton2017} and autotetraploids (\cref{eq:rectet}) to the mixed-ploidy
case. 
Denoting the parents of individual $i$ by $k$ and $l$, the recursion for the
inbreeding coefficients in the mixed-ploidy case becomes
\begin{align}
    F_i &= \Phi_{kl} & \text{if } & c_i = 2 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_k^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 2, g_l = 1 \nonumber \\ 
    F_i &= \frac{1}{3}\left(F_l^\ast + 2\Phi_{kl}\right) & \text{if } 
        & c_i = 3, g_k = 1, g_l = 2 \nonumber \\ 
    F_i &= \frac1 6 (F_k^\ast + F_l^\ast + 4\Phi_{kl}) & \text{if } & c_i = 4
    \label{eq:coanc}
\end{align}
where $F_k^\ast = \alpha_{c_k} + (1-\alpha_{c_k})F_k$, as in the
autotetraploid model.
The recursion for the coancestry coefficients in \cref{eq:coanc} is given by
\begin{align}
    \Phi_{ii} &= \frac{1}{c_{i}} \left(1 + (c_i-1)F_i\right) \nonumber \\
    \Phi_{ij} &= \sum_k \sum_l P_{ik}P_{jl} \Phi_{kl} & i \ne j 
    \label{eq:coancestry}
\end{align}
where the sums are over individuals in the parental population, and where
$P_{ik} \in \{0, \frac1 3, \frac1 2, \frac2 3, 1\}$ is the probability that a
gene copy in $i$ is derived from parent $k$.

@@

Consider a single locus in a population of $k$-ploids, and let $X_1,\dots,X_k$
denote a random genotype at this locus, where $X_i$ is the additive genetic
value of the allele on homolog $i$.
The variance in gametic values $Y$ produced by some meiotic process can be
decomposed as 
\begin{equation}
  \Var[Y] = \underbrace{\Ex\left[\Var[Y|X_1, \dots, X_k]\right]}_{\text{segregation variance}} 
    + \underbrace{\Var\left[\Ex[Y|X_1, \dots, X_k]\right]}_{\text{population
variation}}
\label{eq:eve}
\end{equation}
We use this relationship to derive the segregation variance associated with
haploid and diploid gamete production in the three cytotypes of the
mixed-ploidy population.

We write $V_{S(k,l)}$ for the segregation variance associated with a
$k$-ploid individual producing an $l$-ploid gamete, and we express this in
terms of the variance $\Vx_{k}$ associated with a haploid genome in a
(hypothetical) $k$-ploid reference population at HWLE.
Using the scaling assumptions outlined in the main text, i.e. $\Vx_{k} =
\beta_k^2 V$, we can then express all segregation variances in terms of the
diploid segregation variance in the reference population $V$.

As an example, consider ordinary diploid meiosis at a single locus, where a
diploid produces a haploid gamete. $Y$ is a random haploid gamete sampled from the
population, and $\Var[Y] = \Var[X] = v_x$, where $X$ is a a randomly sampled
gene from the population and $v_x$ is the variance in additive genetic values
at the locus.
We have
  $$\Var[\Ex[Y|X_1,X_2]] = \Var\left[\frac{X_1 + X_2}{2}\right] = \frac{v_x}{2}$$
Using \cref{eq:eve}, we can then find the segregation variance
  $$\Ex[\Var[Y|X_1,X_2]] = v_x - \frac{v_x}{2} = \frac{v_x}{2}$$
Summing over $L$ independent loci, we find the gametic segregation variance in
diploids producing haploid gametes (which is halve the zygotic segregation
variance $V$, by definition) as
  $$V_{S(2,1)} = \frac{V}{2} = \sum_{i=1}^L \frac{v_{x,i}}{2} = \frac{\Vx_{2}}{2}$$
and hence $V=\Vx_{2}$, where $\Vx_{2}$ is the genetic variance associated with
a haploid genome in the diploid cytotype, as defined in the main text.
This will hold in the infinitsimal limit, where $L$ becomes very large and
$v_x$ smaller and smaller.

@@

The mechanisms of unreduced gamete formation do not necessarily lead to a
faithful transmission of the complete diploid genome.
Unreduced gametes are formed in two ways, depending on the meiotic abberation
that leads to their origin: (1) first division restitution (FDR) of (2) second
division restitution (SDR) \citep{bretagnolle1995,storme2013}.
Consider a locus in a diploid with two distinct genes $A$ and $a$. Assume
recombination between the centromere and the locus happens with probability $c$
and that conditional on unreduced gamete formation, formation is due to FDR
with probability $f$ while it is due to SDR with probability $1-f$.
The different unreduced gametes that are formed are represented schematically
in \cref{fig:udip}.
Writing the genotype at a locus in the diploid parent as $X_1X_2$, with allelic
effects $X_1$ and $X_2$, the genotypic value of an unreduced gamete at this
locus will be
$$Y = \begin{cases}
    2X_1 & \text{w.pr.  } \alpha_2/2 \\
    2X_2 & \text{w.pr.  } \alpha_2/2 \\
    X_1 + X_2 & \text{w.pr.  } 1-\alpha_2 
    \end{cases}$$
where $\alpha_2 = 1 - f - c + \frac{3}{2}cf$ is the probability that two copies
of the same gene end up in a diploid gamete produced by a diploid individual
(see the diagram in \cref{fig:udip}). 
Conditional on the latter event, we get
\begin{align*}
\Ex[\Var[Y|X_1,X_2]] &= \Var[Y] - \Var[\Ex[Y|X_1,X_2]] \\
    &= \Var[2X] - \Var\left[\frac{1}{2}(2X_1 + 2X_2)\right] \\
    &= 2v_x
\end{align*}
Conditioning on the complementary event, all gametes have genetic value
$Y = X_1+X_2$, so that the segregation variance is 0.
Summing across independent loci, we have $V_{S(2,2)} = 2\alpha_2 \Vx_{2}$.
